---
title: "Lab 7.8 - Non-linear Modeling"
author: "Barret Miller"
date: "February 27, 2016"
output: html_document
---
### 7.8.1 Polynomial Regression and Step Functions
```{r}
library(ISLR)
attach(Wage)

fit=lm(wage~poly(age,4),data=Wage)
coef(summary(fit))
```

We can also use poly() to obtain age, age^2, age^3 and age^4 directly, if we prefer. We can do this by using the raw=TRUE argument to the poly() function. Later we see that this does not affect the model in a meaningful way—though the choice of basis clearly affects the coefficient estimates, it does not affect the fitted values obtained.
```{r}
fit2=lm(wage~poly(age,4,raw=T),data=Wage)
coef(summary(fit2))
```

This simply creates the polynomial basis functions on the fly, taking care to protect terms like age^2 via the wrapper function I() (the ^ symbol has a special meaning in formulas).
```{r}
fit2a=lm(wage~age+I(age^2)+I(age^3)+I(age^4),data=Wage)
coef(fit2a)
```

This does the same more compactly, using the cbind() function for building a matrix from a collection of vectors; any function call such as cbind() inside a formula also serves as a wrapper.
```{r}
fit2b=lm(wage~cbind(age,age^2,age^3,age^4),data=Wage)

agelims=range(age)
agelims
age.grid=seq(from=agelims[1],to=agelims[2])
age.grid

preds=predict(fit,newdata=list(age=age.grid),se=T)

se.bands=cbind(preds$fit + 2*preds$se.fit, preds$fit - 2*preds$se.fit)

par(mfrow=c(1,2),mar=c(4.5,4.5,1,1),oma=c(0,0,4,0))
plot(age,wage,xlim=agelims,cex=.5,col="darkgrey")

title("Degree-4 Polynomial",outer=T)
lines(age.grid,preds$fit,lwd=2,col="blue")
matlines(age.grid,se.bands,lwd=1,col="blue",lty=3)

preds2=predict(fit2,newdata=list(age=age.grid),se=T)
max(abs(preds$fit-preds2$fit))
```

In performing a polynomial regression we must decide on the degree of the polynomial to use. One way to do this is by using hypothesis tests. We now fit models ranging from linear to a degree-5 polynomial and seek to determine the simplest model which is sufficient to explain the relationship between wage and age. We use the anova() function, which performs an analysis of variance (ANOVA, using an F-test) in order to test the null hypothesis that a model M1 is sufficient to explain the data against the alternative hypothesis that a more complex model M2 is required. In order to use the anova() function, M1 and M2 must be nested models: the predictors in M1 must be a subset of the predictors in M2. In this case, we fit five different models and sequentially compare the simpler model to the more complex model.
```{r}
fit.1=lm(wage~age,data=Wage)
fit.2=lm(wage~poly(age,2),data=Wage)
fit.3=lm(wage~poly(age,3),data=Wage)
fit.4=lm(wage~poly(age,4),data=Wage)
fit.5=lm(wage~poly(age,5),data=Wage)
anova(fit.1,fit.2,fit.3,fit.4,fit.5)

```

In this case, instead of using the anova() function, we could have obtained these p-values more succinctly by exploiting the fact that poly() creates orthogonal polynomials. Notice that the p-values are the same, and in fact the square of the t-statistics are equal to the F-statistics from the anova() function.
```{r}
coef(summary(fit.5))
```

However, the ANOVA method works whether or not we used orthogonal polynomials; it also works when we have other terms in the model as well. For example, we can use anova() to compare these three models:
```{r}
fit.1=lm(wage~education+age,data=Wage)
fit.2=lm(wage~education+poly(age,2),data=Wage)
fit.3=lm(wage~education+poly(age,3),data=Wage)
anova(fit.1,fit.2,fit.3)
```

As an alternative to using hypothesis tests and ANOVA, we could choose the polynomial degree using cross-validation, as discussed in Chapter 5.
Next we consider the task of predicting whether an individual earns more than $250,000 per year. We proceed much as before, except that first we create the appropriate response vector, and then apply the glm() function using family="binomial" in order to fit a polynomial logistic regression model.
Note that we again use the wrapper I() to create this binary response variable on the fly. The expression wage>250 evaluates to a logical variable containing TRUEs and FALSEs, which glm() coerces to binary by setting the TRUEs to 1 and the FALSEs to 0.
```{r}
fit=glm(I(wage>250)~poly(age,4),data=Wage,family=binomial)
preds=predict(fit,newdata=list(age=age.grid),se=T)
```

The default prediction type for a glm() model is type="link", which is what we use here. This means we get predictions for the logit and the predictions given are of the form Xβˆ. The standard errors given are also of this form. In order to obtain confidence intervals for Pr(Y = 1|X), we must transform:
```{r}
pfit=exp(preds$fit)/(1+exp(preds$fit))
se.bands.logit = cbind(preds$fit+2*preds$se.fit, preds$fit-2*preds$se.fit)
se.bands = exp(se.bands.logit)/(1+exp(se.bands.logit))
```

Note that we could have directly computed the probabilities by selecting the type="response" option in the predict() function. However, the corresponding confidence intervals would not have been sen- sible because we would end up with negative probabilities!
Below, we have drawn the age values corresponding to the observations with wage values above 250 as gray marks on the top of the plot, and those with wage values below 250 are shown as gray marks on the bottom of the plot. We used the jitter() function to jitter the age values a bit so that observations with the same age value do not cover each other up. This is often called a rug plot.
```{r}
plot(age,I(wage>250),xlim=agelims,type="n",ylim=c(0,.2))
points(jitter(age), I((wage>250)/5), cex=.5, pch="|", col="darkgrey")
lines(age.grid,pfit,lwd=2,col="blue")
matlines(age.grid,se.bands,lwd=1,col="blue",lty=3)
```

In order to fit a step function, as discussed in Section 7.2, we use the cut() function. Here cut() automatically picked the cutpoints at 33.5, 49, and 64.5 years of age. We could also have specified our own cutpoints directly using the breaks option. The function cut() returns an ordered categorical variable; the lm() function then creates a set of dummy variables for use in the re- gression. The age<33.5 category is left out, so the intercept coefficient of $94,160 can be interpreted as the average salary for those under 33.5 years of age, and the other coefficients can be interpreted as the average addi- tional salary for those in the other age groups. We can produce predictions and plots just as we did in the case of the polynomial fit.
```{r}
table(cut(age,4))
fit=lm(wage~cut(age,4),data=Wage)
coef(summary(fit))
```

### 7.8.2 Splines
In order to fit regression splines in R, we use the splines library. In Section
7.4, we saw that regression splines can be fit by constructing an appropriate matrix of basis functions. The bs() function generates the entire matrix of bs() basis functions for splines with the specified set of knots. By default, cubic splines are produced. Fitting wage to age using a regression spline is simple.
Here we have prespecified knots at ages 25, 40, and 60. This produces a spline with six basis functions. (Recall that a cubic spline with three knots has seven degrees of freedom; these degrees of freedom are used up by an intercept, plus six basis functions; one for every level of the cubic polynomial, and one for each truncated power function for each knot.) If you read the documentation for bs, the default is to not model an intercept, so that's why the matrix has 6 terms, instead of 7. We could also use the df option to produce a spline with knots at uniform quantiles of the data. In this case R chooses knots at ages 33.8, 42.0, and 51.0, which correspond to the 25th, 50th, and 75th percentiles of age. The function bs() also has a degree argument, so we can fit splines of any degree, rather than the default degree of 3 (which yields a cubic spline).In order to instead fit a natural spline, we use the ns() function. Here ns() we fit a natural spline with four degrees of freedom. As with the bs() function, we could instead specify the knots directly using the knots option.
```{r}
library(splines)
fit=lm(wage~bs(age,knots=c(25,40,60)),data=Wage)
pred=predict(fit,newdata=list(age=age.grid),se=T)
plot.new()
plot(age,wage,col="gray")
lines(age.grid,pred$fit,lwd=2)
lines(age.grid,pred$fit+2*pred$se.fit,lty="dashed")
lines(age.grid,pred$fit-2*pred$se.fit,lty="dashed")
dim(bs(age,knots=c(25,40,60)))
dim(bs(age,df=6))
attr(bs(age,df=6),"knots")
fit2=lm(wage~ns(age,df=4),data=Wage)
pred2=predict(fit2,newdata=list(age=age.grid),se=T)
lines(age.grid,pred2$fit,col="red",lwd=2)
```

In order to fit a smoothing spline, we use the smooth.spline() function. smooth. Figure 7.8 was produced with the following code.
Notice that in the first call to smooth.spline(), we specified df=16. The function then determines which value of λ leads to 16 degrees of freedom. In the second call to smooth.spline(), we select the smoothness level by cross- validation; this results in a value of λ that yields 6.8 degrees of freedom.
```{r}
plot.new()
plot(age,wage,xlim=agelims,cex=.5,col="darkgrey")
title("Smoothing Spline")
fit=smooth.spline(age,wage,df=16)
fit2=smooth.spline(age,wage,cv=TRUE)
fit2$df
lines(fit,col="red",lwd=2)
lines(fit2,col="blue",lwd=2)
legend("topright",legend=c("16 DF", "6.8 DF"), col=c("red","blue"), lty=1, lwd=2, cex=.8)
```

In order to perform local regression, we use the loess() function.
Here we have performed local linear regression using spans of 0.2 and 0.5: that is, each neighborhood consists of 20 % or 50 % of the observations. The larger the span, the smoother the fit. The locfit library can also be used for fitting local regression models in R.
```{r}
plot.new()
par(mfrow=c(1,1))
plot(age,wage,xlim=agelims,cex=.5,col="darkgrey")
title("Local Regression")
fit=loess(wage~age,span=.2,data=Wage)
fit2=loess(wage~age,span=.5,data=Wage)
lines(age.grid,predict(fit,data.frame(age=age.grid)),col="red",lwd=2)
lines(age.grid,predict(fit2,data.frame(age=age.grid)),col="blue",lwd=2)
legend("topright",legend=c("Span=0.2", "Span=0.5"), col=c("red","blue"),lty=1,lwd=2,cex=.8)
```

### 7.8.3 GAMs
We now fit a GAM to predict wage using natural spline functions of year and age, treating education as a qualitative predictor, as in (7.16). Since this is just a big linear regression model using an appropriate choice of basis functions, we can simply do this using the lm() function.
```{r}
gam1 = lm(wage~ns(year,4)+ns(age,5)+education,data=Wage)
```

We now fit the model (7.16) using smoothing splines rather than natural splines. In order to fit more general sorts of GAMs, using smoothing splines or other components that cannot be expressed in terms of basis functions and then fit using least squares regression, we will need to use the gam library in R.
The s() function, which is part of the gam library, is used to indicate that we would like to use a smoothing spline. We specify that the function of
year should have 4 degrees of freedom, and that the function of age will
have 5 degrees of freedom. Since education is qualitative, we leave it as is,
and it is converted into four dummy variables. We use the gam() function in order to fit a GAM using these components. All of the terms in (7.16) are
fit simultaneously, taking each other into account to explain the response.
The generic plot() function recognizes that gam.m3 is an object of class gam, andinvokestheappropriateplot.gam()method.Conveniently,eventhough gam1 is not of class gam but rather of class lm, we can still use plot.gam()
on it. Figure 7.11 was produced using the following expression
```{r}
library(gam)
gam.m3=gam(wage~s(year,4)+s(age,5)+education,data=Wage)
plot.new()
par(mfrow=c(1,3))
plot(gam.m3,se=TRUE,col="blue")
plot.gam(gam1,se=TRUE,col="red")
```

In these plots, the function of year looks rather linear. We can perform a series of ANOVA tests in order to determine which of these three models is best: a GAM that excludes year (M1), a GAM that uses a linear function of year (M2), or a GAM that uses a spline function of year (M3).
We find that there is compelling evidence that a GAM with a linear func- tion of year is better than a GAM that does not include year at all (p-value = 0.00014). However, there is no evidence that a non-linear func- tion of year is needed (p-value = 0.349). In other words, based on the results of this ANOVA, M2 is preferred.
```{r}
gam.m1=gam(wage~s(age,5)+education,data=Wage)
gam.m2=gam(wage~year+s(age,5)+education,data=Wage)
anova(gam.m1,gam.m2,gam.m3)
```

The p-values for year and age correspond to a null hypothesis of a linear relationship versus the alternative of a non-linear relationship. The large p-value for year in the "Anova for Nonparametric Effects" section of the output reinforces our conclusion from the ANOVA test that a linear function is adequate for this term. However, there is very clear evidence that a non-linear term is required for age.
```{r}
summary(gam.m3)
```

We can make predictions from gam objects, just like from lm objects, using the predict() method for the class gam. Here we make predictions on the training set.
```{r}
preds=predict(gam.m2,newdata=Wage)
```

We can also use local regression fits as building blocks in a GAM, using
the lo() function. Here we have used local regression for the age term, with a span of 0.7
```{r}
gam.lo=gam(wage~s(year,df=4)+lo(age,span=.7)+education,data=Wage)
plot.new()
plot.gam(gam.lo,se=TRUE,col="green")
```

We can also use the lo() function to create interactions before calling the gam() function. Below fits a two-term model, in which the first term is an interaction between year and age, fit by a local regression surface. We can plot the resulting two-dimensional surface if we first install the akima package.
```{r}
gam.lo.i=gam(wage~lo(year,age,span=.5)+education,data=Wage)
library(akima)
plot.new()
plot(gam.lo.i)
```

In order to fit a logistic regression GAM, we once again use the I() function in constructing the binary response variable, and set family=binomial.
```{r}
gam.lr=gam(I(wage>250)~year+s(age,df=5)+education,family=binomial,data=Wage)
plot.new()
par(mfrow=c(1,3))
plot(gam.lr,se=T,col="green")
```

It is easy to see that there are no high earners in the "\<HS"" category. Hence, we fit a logistic regression GAM using all but this category. This provides more sensible results.
```{r}
table(education,I(wage>250))
gam.lr.s=gam(I(wage>250)~year+s(age,df=5)+education,family=binomial,data=Wage
             ,subset = (education!="1. < HS Grad"))
plot.new()
plot(gam.lr.s,se=T,col="green")
```