---
title: "Homework - Chapter 8"
author: "Barret Miller"
date: "March 7, 2016"
output: html_document
---

### 8.4.7
```{r}
library(MASS)
library(randomForest)
attach(Boston)
data(Boston)

set.seed(1)
train = sample(nrow(Boston),nrow(Boston)/2)

ntrees = seq(10,500,by=10)
nvars = c(ncol(Boston)-1, ceiling((ncol(Boston)-1)/2), ceiling(sqrt(ncol(Boston)-1)))

testError = matrix(NA,ncol=3,nrow=50)
for(i in 1:length(nvars)){
   nv = nvars[i]
   for(j in 1:length(ntrees)){
      nt = ntrees[j]
      rf = randomForest(medv~.,data=Boston,subset=train,mtry=nv,ntree=nt)
      yhat = predict(rf,newdata=Boston[-train,])
      #print(paste(i,j))
      testError[j,i] = mean((yhat-Boston[-train,"medv"])^2)
   }
}

matplot(ntrees,testError,col=c("red","blue","green"),type="l",lwd=c(2,2,3))
legend("topright",legend=c("m=p", "m=p/2", paste0("m=",expression(sqrt(p)))), col=c("red","blue","green"),lty = c(1,2,2))
```

An alternative approach would be to just use the built-in features of the randomForest library, which allow you to pass in a test set, and it will automatically return the test MSE vector. This returns a value for every number of trees incrementing by 1, instead of by 10s as I did. Also it is much faster, so it must have some optimizations. I pulled this code straight from: [https://rpubs.com/ppaquay/65564](https://rpubs.com/ppaquay/65564)
From the documentation for randomForest:

> If xtest is given, prediction of the test set is done “in place” as the trees are grown.

```{r}
set.seed(1)
train <- sample(1:nrow(Boston), nrow(Boston) / 2)
Boston.train <- Boston[train, -14]
Boston.test <- Boston[-train, -14]
Y.train <- Boston[train, 14]
Y.test <- Boston[-train, 14]
rf.boston1 <- randomForest(Boston.train, y = Y.train, xtest = Boston.test, ytest = Y.test, mtry = ncol(Boston) - 1, ntree = 500)
rf.boston2 <- randomForest(Boston.train, y = Y.train, xtest = Boston.test, ytest = Y.test, mtry = (ncol(Boston) - 1) / 2, ntree = 500)
rf.boston3 <- randomForest(Boston.train, y = Y.train, xtest = Boston.test, ytest = Y.test, mtry = sqrt(ncol(Boston) - 1), ntree = 500)
plot(1:500, rf.boston1$test$mse, col = "green", type = "l", xlab = "Number of Trees", ylab = "Test MSE", ylim = c(10, 19))
lines(1:500, rf.boston2$test$mse, col = "red", type = "l")
lines(1:500, rf.boston3$test$mse, col = "blue", type = "l")
legend("topright", c("m = p", "m = p/2", "m = sqrt(p)"), col = c("green", "red", "blue"), cex = 1, lty = 1)
```