---
title: "Lab 8.3"
author: "Barret Miller"
date: "March 6, 2016"
output: html_document
---

### 8.3.1 Fitting Classification Trees

```{r}
library(tree)
library(ISLR)
attach(Carseats)

High=ifelse(Sales<=8,"No","Yes")
Carseats = data.frame(Carseats,High)
```

We now use the tree() function to fit a classification tree in order to predict High using all variables but Sales. The syntax of the tree() function is quite similar to that of the lm() function. The summary() function lists the variables that are used as internal nodes in the tree, the number of terminal nodes, and the (training) error rate. The explanation of the "Residual mean deviance" is on page 325 (PDF page 338) of the ISL book.

Deviance (the numerator: 170.7 in this case):

-2 times the sum over all the partitions, and for each partition, sum the following for each class: the number of observations in the class in that partition, times log of the fraction or proportion of that class in that partition.

The denominator is n - |T0|, which is the number of observations, n, minus the number of terminal nodes. 
```{r}
tree.carseats = tree(High~.-Sales,Carseats)
summary(tree.carseats)
```

We use the plot() function to display the tree struc- ture, and the text() function to display the node labels. The argument pretty=0 instructs R to include the category names for any qualitative pre- dictors, rather than simply displaying a letter for each category.
```{r}
plot(tree.carseats)
text(tree.carseats,pretty=0)
```

If we just type the name of the tree object, R prints output corresponding to each branch of the tree. R displays the split criterion (e.g. Price<92.5), the number of observations in that branch, the deviance, the overall prediction for the branch (Yes or No), and the fraction of observations in that branch that take on values of Yes and No. Branches that lead to terminal nodes are indicated using asterisks. Notice that those terminal nodes with completely pure fractions of Yes's and No's have 0 deviance. 
```{r}
tree.carseats
```

In order to properly evaluate the performance of a classification tree on these data, we must estimate the test error rather than simply computing the training error. We split the observations into a training set and a test set, build the tree using the training set, and evaluate its performance on the test data. The predict() function can be used for this purpose. In the case of a classification tree, the argument type="class" instructs R to return the actual class prediction. This approach leads to correct predictions for around 71.5 % of the locations in the test data set.
```{r}
set.seed(2)
train=sample(1:nrow(Carseats),200)
Carseats.test = Carseats[-train,]
High.test=High[-train]

tree.carseats=tree(High~.-Sales,Carseats,subset=train)
tree.pred=predict(tree.carseats,Carseats.test,type="class")
table(tree.pred,High.test)
(86+57)/200
```

Next, we consider whether pruning the tree might lead to improved results. The function cv.tree() performs cross-validation in order to determine the optimal level of tree complexity; cost complexity pruning is used in order to select a sequence of trees for consideration. We use the argument FUN=prune.misclass in order to indicate that we want the classification error rate to guide the cross-validation and pruning process, rather than the default for the cv.tree() function, which is deviance. The cv.tree() function reports the number of terminal nodes of each tree con- sidered (size) as well as the corresponding error rate and the value of the cost-complexity parameter used (k, which corresponds to α in (8.4)).
```{r}
set.seed(3)
cv.carseats = cv.tree(tree.carseats,FUN=prune.misclass)
names(cv.carseats)
cv.carseats
```

Note that, despite the name, dev corresponds to the cross-validation error rate in this instance. The tree with 9 terminal nodes results in the lowest cross-validation error rate, with 50 cross-validation errors. We plot the error rate as a function of both size and k.
```{r}
par(mfrow=c(1,2))
plot(cv.carseats$size, cv.carseats$dev,type="b")
plot(cv.carseats$k, cv.carseats$dev, type="b")
```

We now apply the prune.misclass() function in order to prune the tree to obtain the nine-node tree.
```{r}
prune.carseats = prune.misclass(tree.carseats,best=9)
plot(prune.carseats)
text(prune.carseats,pretty=0)
```

How well does this pruned tree perform on the test data set? Once again, we apply the predict() function.
```{r}
tree.pred=predict(prune.carseats,Carseats.test,type="class")
table(tree.pred,High.test)
(94+60)/200
mean(tree.pred == High.test)
```

If we increase the value of best, we obtain a larger pruned tree with lower classification accuracy
```{r}
prune.carseats = prune.misclass(tree.carseats,best=15)
plot(prune.carseats)
text(prune.carseats,pretty=0)
tree.pred = predict(prune.carseats, Carseats.test, type="class")
table(tree.pred,High.test)
mean(tree.pred == High.test)
```

### 8.3.2 Fitting Regression Trees
Here we fit a regression tree to the Boston data set. First, we create a training set, and fit the tree to the training data.

```{r}
library(MASS)
set.seed(1)
train = sample(1:nrow(Boston), nrow(Boston)/2)
tree.boston = tree(medv~.,Boston,subset=train)
summary(tree.boston)
```

Notice that the output of summary() indicates that only three of the variables have been used in constructing the tree. In the context of a regression tree, the deviance (numerator) is simply the sum of squared errors for the tree. The denominator is again n minus |T0|, or the number of terminal nodes, so we get 253-8=245. So, the average RSS per terminal node. We now plot the tree.
```{r}
plot(tree.boston)
text(tree.boston,pretty=0)
```

The variable lstat measures the percentage of individuals with lower socioeconomic status. The tree indicates that lower values of lstat cor- respond to more expensive houses. The tree predicts a median house price of $46, 400 for larger homes in suburbs in which residents have high socioe- conomic status (rm>=7.437 and lstat<9.715). 
Now we use the cv.tree() function to see whether pruning the tree will improve performance.
```{r}
cv.boston=cv.tree(tree.boston)
plot(cv.boston$size,cv.boston$dev,type='b')
```

In this case, the most complex tree is selected by cross-validation. How- ever, if we wish to prune the tree, we could do so as follows, using the prune.tree() function:
```{r}
prune.boston = prune.tree(tree.boston,best=5)
plot(prune.boston)
text(prune.boston,pretty=0)
```

In keeping with the cross-validation results, we use the unpruned tree to make predictions on the test set.
```{r}
yhat=predict(tree.boston,newdata=Boston[-train,])
boston.test=Boston[-train,"medv"]
plot(yhat,boston.test)
abline(0,1)
MSE=mean((yhat-boston.test)^2)
RMSE=sqrt(MSE)
MSE
RMSE
```

In other words, the test set MSE associated with the regression tree is 25.05. The square root of the MSE is therefore around 5.005, indicating that this model leads to test predictions that are within around $5, 005 of the true median home value for the suburb.

### 8.3.3 Bagging and Random Forests
Here we apply bagging and random forests to the Boston data, using the randomForest package in R. The exact results obtained in this section may depend on the version of R and the version of the randomForest package installed on your computer. Recall that bagging is simply a special case of a random forest with m = p. Therefore, the randomForest() function can be used to perform both random forests and bagging. We perform bagging as follows:
```{r}
library(randomForest)
set.seed(1)
bag.boston=randomForest(medv~.,data=Boston,subset=train,mtry=13,importance=TRUE)
bag.boston
```

The argument mtry=13 indicates that all 13 predictors should be considered for each split of the tree—in other words, that bagging should be done. How well does this bagged model perform on the test set?
```{r}
yhat.bag = predict(bag.boston,newdata=Boston[-train,])
plot(yhat.bag,boston.test)
abline(0,1)
mean((yhat.bag-boston.test)^2)
```

The test set MSE associated with the bagged regression tree is 13.47, almost half that obtained using an optimally-pruned single tree. We could change the number of trees grown by randomForest() using the ntree argument:
```{r}
bag.boston=randomForest(medv~.,data=Boston,subset=train,mtry=13,ntree=25)
yhat.bag = predict(bag.boston,newdata=Boston[-train,])
mean((yhat.bag-boston.test)^2)
```

Growing a random forest proceeds in exactly the same way, except that we use a smaller value of the mtry argument. By default, randomForest() uses p/3 variables when building a random forest of regression trees, and √p variables when building a random forest of classification trees. Here we use mtry = 6.
```{r}
set.seed(1)
rf.boston=randomForest(medv~.,data=Boston,subset=train,mtry=6,importance=TRUE)
yhat.rf=predict(rf.boston,newdata=Boston[-train,])
mean((yhat.rf-boston.test)^2)
```

The test set MSE is 11.2; this indicates that random forests yielded an improvement over bagging in this case.
Using the importance() function, we can view the importance of each importance() variable.
```{r}
importance(rf.boston)
```

Two measures of variable importance are reported. The former is based upon the mean decrease of accuracy in predictions on the out of bag samples when a given variable is excluded from the model. The latter is a measure of the total decrease in node impurity that results from splits over that variable, averaged over all trees (this was plotted in Figure 8.9). In the case of regression trees, the node impurity is measured by the training RSS, and for classification trees by the deviance. Plots of these importance measures can be produced using the varImpPlot() function.
```{r}
varImpPlot(rf.boston)
```

The results indicate that across all of the trees considered in the random forest, the wealth level of the community (lstat) and the house size (rm) are by far the two most important variables.

### 8.3.4 Boosting
Here we use the gbm package,and within it the gbm() function, to fit boosted regression trees to the Boston data set. We run gbm() with the option distribution="gaussian" since this is a regression problem; if it were a bi- nary classification problem, we would use distribution="bernoulli". The argument n.trees=5000 indicates that we want 5000 trees, and the option interaction.depth=4 limits the depth of each tree.
```{r}
library(gbm)
set.seed(1)
boost.boston=gbm(medv~.,data=Boston[train,],distribution="gaussian",n.trees=5000,interaction.depth=4)
```

The summary() function produces a relative influence plot and also outputs the relative influence statistics.
```{r}
summary(boost.boston)
```

We see that lstat and rm are by far the most important variables. We can also produce partial dependence plots for these two variables. These plots illustrate the marginal effect of the selected variables on the response after integrating out the other variables. In this case, as we might expect, median house prices are increasing with rm and decreasing with lstat.
```{r}
par(mfrow=c(1,2))
plot(boost.boston,i="rm")
plot(boost.boston,i="lstat")
```

We now use the boosted model to predict medv on the test set:
```{r}
yhat.boost=predict(boost.boston,newdata=Boston[-train,],n.trees=5000)
mean((yhat.boost-boston.test)^2)
```

The test MSE obtained is 11.8; similar to the test MSE for random forests and superior to that for bagging. If we want to, we can perform boosting with a different value of the shrinkage parameter λ in (8.10). The default value is 0.001, but this is easily modified. Here we take λ = 0.2
```{r}
boost.boston=gbm(medv~.,data=Boston[train,],distribution = "gaussian",n.trees=5000,interaction.depth = 4,shrinkage=0.2,verbose = F)
yhat.boost=predict(boost.boston,newdata=Boston[-train,],n.trees=5000)
mean((yhat.boost-boston.test)^2)
```

In this case, using λ = 0.2 leads to a slightly lower test MSE than λ = 0.001.